---
title: "AVRcap1b_v1"
author: "AmirAli Toghani"
date: "2025-02-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(readxl)
library(Biostrings)
library(ggtree)
library(reshape2)
library(svglite)
library(entropy)
library(ggseqlogo)
library(plotly)
```

#### 1. Import the BLAST output for AVRcap1b:
```{r}
# import the sequences and process the names
AVRcap1b_psiblast_seq <- readAAStringSet("supplementary/AVRcap1b_psiblast.fasta")

AVRcap1b_psiblast_seq_ID <- AVRcap1b_psiblast_seq@ranges@NAMES %>% as.data.frame() %>% setNames("ID")

AVRcap1b_psiblast_seq_ID$ID <- gsub(" .*", "", AVRcap1b_psiblast_seq_ID$ID)

AVRcap1b_psiblast_seq@ranges@NAMES <- AVRcap1b_psiblast_seq_ID$ID


# import the metadata file
AVRcap1b_psiblast_metadata <- read_delim("supplementary/AVRcap1b_psiblast.csv")

AVRcap1b_psiblast_metadata$Accession <- gsub('.*,"(.*?)"\\).*', '\\1', AVRcap1b_psiblast_metadata$Accession)

AVRcap1b_psiblast_metadata <- AVRcap1b_psiblast_metadata %>% 
  separate(`Scientific Name`, into = c("genus","species","strain"), sep = " ", fill = "right", extra = "merge")


AVRcap1b_psiblast_metadata <- AVRcap1b_psiblast_metadata %>% 
  mutate(ID = paste(genus, species, Accession, sep = "_")) %>%
  mutate(ID = gsub(" ", "_", ID))

AVRcap1b_psiblast_metadata <- AVRcap1b_psiblast_metadata[match(AVRcap1b_psiblast_seq@ranges@NAMES, AVRcap1b_psiblast_metadata$Accession),]

AVRcap1b_psiblast_metadata$organism <- trimws(paste(AVRcap1b_psiblast_metadata$genus, AVRcap1b_psiblast_metadata$species," "))


# rename the BLAST set with new IDs
AVRcap1b_psiblast_seq@ranges@NAMES <- AVRcap1b_psiblast_metadata$ID

```


#### Import the reference sequences:
```{r}
AVRcap1b_ref <- readAAStringSet("supplementary/AVRcap1b_ortho.fasta")

# make short description dataframe
AVRcap1b_ref_metadata <- data_frame(ID = AVRcap1b_ref@ranges@NAMES, 
                                    genus = "Phytophthora", 
                                    species = c("andina","infestans","ipomoeae","mirabilis"),
                                    clade = "1c")

AVRcap1b_ref_metadata$organism <- trimws(paste(AVRcap1b_ref_metadata$genus, AVRcap1b_ref_metadata$species," "))

```


#### Tidy up the psiBLAST set:
```{r}
AVRcap1b_psiblast_seq_filtered <- AVRcap1b_psiblast_seq[AVRcap1b_psiblast_seq@ranges@width > 500]
AVRcap1b_psiblast_seq_filtered <- AVRcap1b_psiblast_seq_filtered[AVRcap1b_psiblast_seq_filtered@ranges@width < 800]

AVRcap1b_psiblast_metadata_filtered <- AVRcap1b_psiblast_metadata[match(AVRcap1b_psiblast_seq_filtered@ranges@NAMES, AVRcap1b_psiblast_metadata$ID),]
write_csv(AVRcap1b_psiblast_metadata_filtered, "/path/to/output/Table_SX1.csv")

```



#### Add the references to the BLAST set and export for phylogenetic analysis:
```{r}
AVRcap1b_all_ref <- c(AVRcap1b_psiblast_seq_filtered, AVRcap1b_ref)

writeXStringSet(AVRcap1b_all_ref, "/path/to/output/AVRcap1b_psiblast_all_ref.fasta")
```


#### Calculate the frequencies of sequences per species:
```{r}
# import the subclades
AVRcap1b_like_superclade_tree <- read.tree("supplementary/AVRcap1b_like_superclade.tree")

AVRcap1b_like_superclade_metadata <- AVRcap1b_psiblast_metadata_filtered[AVRcap1b_psiblast_metadata_filtered$ID %in% AVRcap1b_like_superclade_tree$tip.label,]

AVRcap1b_like_superclade_ref_seq <- AVRcap1b_all_ref[AVRcap1b_like_superclade_tree$tip.label]
AVRcap1b_like_superclade_seq <- AVRcap1b_all_ref[AVRcap1b_like_superclade_metadata$ID]

writeXStringSet(AVRcap1b_like_superclade_ref_seq, "/path/to/output/AVRcap1b_like_superclade_ref.fasta")


PSR2_like_superclade_tree <- read.tree("supplementary/PSR2_like_superclade.tree")

PSR2_like_superclade_metadata <- AVRcap1b_psiblast_metadata_filtered[AVRcap1b_psiblast_metadata_filtered$ID %in% PSR2_like_superclade_tree$tip.label,]



# calculate the frequencies
AVRcap1b_psiblast_freq <- table(AVRcap1b_psiblast_metadata_filtered$organism) %>% as.data.frame() %>% setNames(c("species","total"))

AVRcap1b_like_freq <- table(AVRcap1b_like_superclade_metadata$organism) %>% as.data.frame() %>% setNames(c("species","AVRcap1b-like"))

PSR2_like_freq <- table(PSR2_like_superclade_metadata$organism) %>% as.data.frame() %>% setNames(c("species","PSR2-like"))

species_freq <- Reduce(function(x,y) merge(x,y, by = "species", all = TRUE), 
                       list(AVRcap1b_psiblast_freq, AVRcap1b_like_freq, PSR2_like_freq))

species_freq[is.na(species_freq)] <- 0

# reference frequencies
ref_freq <- data_frame(species = c("Phytophthora andina","Phytophthora ipomoeae","Phytophthora mirabilis"), 
                       total = "1",
                       "AVRcap1b-like" = "1",
                       "PSR2-like" = "0")

species_freq <- rbind(species_freq,ref_freq)

species_list <- read_csv("supplementary/species_list_ext.txt")

species_freq <- species_freq[match(species_list$species, species_freq$species),]

# plot the frequencies
species_freq_melted <- melt(species_freq, id.vars = "species", measure.vars = c("AVRcap1b-like", "PSR2-like"))

species_freq_melted$value <- as.numeric(species_freq_melted$value)

species_freq_melted$species <- factor(species_freq_melted$species, levels = rev(unique(species_freq_melted$species)))
species_freq_melted$variable <- factor(species_freq_melted$variable, levels = rev(unique(species_freq_melted$variable)))

p_species_freq <- ggplot(species_freq_melted, aes(x = species, y = value, fill = variable)) +
  geom_bar(stat = "identity", show.legend = F) +
  coord_flip() +
  labs(x = NULL, y = NULL, fill = NULL) +  # Remove axis and legend titles
  theme_minimal() +  
  theme(
    panel.grid = element_blank(),  # Remove gridlines
    panel.background = element_blank(),  # Remove background
    axis.text.y = element_text(size = 10),  # Keep y-axis labels readable
    axis.text.x = element_text(size = 10, vjust = -1),  # Move x-axis labels to the top
    axis.ticks = element_blank(),  # Remove axis ticks
    axis.line.x = element_line(color = "black"),  # Keep only the x-axis line
    axis.line.y = element_blank(),  # Remove y-axis line
    legend.position = "none"  # Remove legend
    ) +
  scale_fill_manual(values = c("AVRcap1b-like" = "#2A3439", "PSR2-like" = "#BEBFC5"))


ggsave(p_species_freq, width = 3, height = 10, filename = "/path/to/output/species_freq.svg", dpi = "retina", device = "svg")

```


#### Import clade details and generate annotations for iTOL:
```{r}
species_clades <- read_excel("supplementary/species_clades.xlsx")

species_freq <- species_clades %>% left_join(species_freq, by = "species")



AVRcap1b_psiblast_metadata_filtered_annot <- AVRcap1b_psiblast_metadata_filtered %>% left_join(species_clades, join_by(organism == species))


# add subclade colors
AVRcap1b_psiblast_metadata_filtered_annot <- AVRcap1b_psiblast_metadata_filtered_annot %>% mutate(subclade_color = case_when(
  clade == "1a" ~ "#0077b8",
  clade == "1c" ~ "#003958",
  clade == "1d" ~ "#0077b8",
  clade == "2a" ~ "#61b7e7",
  clade == "2b" ~ "#61b7e7",
  clade == "2d" ~ "#61b7e7",
  clade == "3" ~ "#f47527",
  clade == "4" ~ "#febf10",
  clade == "7a" ~ "#d7e594",
  clade == "7b" ~ "#d7e594",
  clade == "7c" ~ "#d7e594",
  clade == "7d" ~ "#d7e594",
  clade == "11" ~ "#8cc864",
  clade == "8c" ~ "#16a64a"))


write_csv(AVRcap1b_psiblast_metadata_filtered_annot[,c(11:15)],"/path/to/output/AVRcap1b_psiblast_annot.csv")

# same for references...clade color
AVRcap1b_ref_metadata_annot <- AVRcap1b_ref_metadata %>% mutate(clade_color = case_when(clade == "1c" ~ "#003958"))

write_csv(AVRcap1b_ref_metadata_annot,"/path/to/output/AVRcap1b_ref_annot.csv")


```




#### Import AVRcap1b clade alignment for entropy:
Two functions for calculating the entropy:
```{r}
# Function to convert a Biostrings/dataframe object to a dataframe and break the sequence into characters
convert_to_dataframe <- function(alignment) {
  # Extract sequences as strings
  sequences <- sapply(alignment, as.character)
  # Split each sequence into individual characters and convert to a matrix
  alignment_matrix <- do.call(rbind, strsplit(sequences, ""))
  # Convert the matrix to a dataframe
  alignment_df <- as.data.frame(alignment_matrix, stringsAsFactors = FALSE)
  # Naming columns as positions
  colnames(alignment_df) <- paste(seq_len(ncol(alignment_df)))
  
  return(alignment_df)
}



# Function to calculate entropy for multiple columns and return as a dataframe
calculate_entropy_df <- function(dataframe) {
  # Function to calculate entropy for a single column
  calculate_column_entropy <- function(column) {
    # Remove any NA values (if there are any)
    column <- na.omit(column)
    # Calculate frequencies
    freq_table <- table(column)
    # Convert to proportions
    proportions <- freq_table / sum(freq_table)
    # Calculate and return entropy
    entropy.empirical(proportions, unit = "log2")
  }
  
  # Ensure that the dataframe is processed column-wise
  entropy_values <- sapply(dataframe, calculate_column_entropy, simplify = FALSE)
  
  # Flatten the list if necessary
  if (is.list(entropy_values)) {
    entropy_values <- unlist(entropy_values)
  }
  
  # Create a dataframe with positions and entropy values
  entropy_df <- data.frame(Position = seq_along(entropy_values), Entropy = entropy_values)
  
  return(entropy_df)
}
```

```{r}
AVrcap1b_clade_aln <- readAAStringSet("supplementary/AVRcap1b_like_superclade_ref_filtered_trimmed.afa")


AVrcap1b_clade_aln_df <- convert_to_dataframe(AVrcap1b_clade_aln)
AVrcap1b_clade_entropy <- calculate_entropy_df(AVrcap1b_clade_aln_df)


```

graphics prep:
```{r}
breakpoints <- c(0, 1.5, 4.32)


theme_custom2 <- theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  legend.position = "none",
  panel.background = element_blank(),
  plot.background = element_blank(),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.title = element_blank(),
  plot.subtitle = element_blank(),
  plot.caption = element_blank(),
  axis.line = element_line(colour = "black", size = 0.25)
)

```

Plot it:
```{r}
AVrcap1b_clade_entropy_p <- ggplot(AVrcap1b_clade_entropy, aes(x = Position, y = Entropy)) +
  geom_line(alpha = 0.5) +
  geom_point(data = subset(AVrcap1b_clade_entropy, Entropy > 1.5), aes(x = Position, y = Entropy), color = "#CC3333", size = 1) +
  scale_color_identity() + # Use the colors as they are
  ylim(c(0,4.32)) +
  xlim(c(0,700)) +
  geom_hline(yintercept = 1.5, linetype = "dotted", color = "#6C6C6C", size = 0.35, alpha = 1) +
  theme_custom2
AVrcap1b_clade_entropy_p

ggplotly(AVrcap1b_clade_entropy_p)
```


#### Import the AVRcap1b clade for phylogenetic analysis:
```{r}
AVRcap1b_clade <- read.tree("supplementary/AVRcap1b_clade.tree")

AVRcap1b_clade_seq <- AVRcap1b_all_ref[AVRcap1b_clade$tip.label]

writeXStringSet(AVRcap1b_clade_seq, "/path/to/output/AVRcap1b_clade.fasta")
```

